Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-24 paper/ms.tex
Remove edits for arxiv posting

\begin{abstract}
    The elemental abundance distribution of stars encodes the history of the gas-phase abundance in the Milky Way. Without a large, unbiased sample of highly precise stellar ages, the exact timing and nature of this history must be \textit{inferred} from the abundances. In the two-dimensional plane of \alphaFe{}-\FeH{}, it is now clear that two separate populations exist -- the low-$\alpha$ and high-$\alpha$ sequences. \edit1{We propose that a brief ($\sim300\,\Myr$) halt in star formation within a narrow metallicity bin can lead to a bimodal \alphaFe{} distribution at that metallicity, assuming a rapidly declining gas phase \alphaFe{}. Using simulations of an idealized setup of a high-$z$ galaxy merger, we show that the merger with the Gaia-Sausage-Enceladus satellite at $z\sim2$ is one possible way to trigger such a gap in the Milky Way. This mechanism may also operate in non-merger scenarios. We predict a $\sim300\,\Myr$ gap in stellar ages at a fixed \FeH{} where the $\alpha$-bimodality is prominent ($\FeH\lesssim-0.2$).}
\end{abstract}

@@

\section{Introduction} \label{sec:intro}
% What leads to distribution of heavy elements in a galaxy
Many elements heavier than hydrogen are produced through nuclear fusion in compact objects such as supernovae, dying low mass stars, and neutron star-neutron star mergers \citep[e.g.][]{2023A&ARv..31....1A}. By necessity, stars inherit the constitutive properties of the gas from which they formed. Moreover, the surface abundance of most elements for most stars do not change over most of their lifetime. \edit1{By analyzing the surface abundances of stars, we can reconstruct the historical gas-phase composition of a galaxy.}

The enrichment of the gas-phase of a galaxy is determined by a complicated combination of physical processes - stellar evolution and supernovae, gas accretion, mergers, gas outflows from stellar and active galactic nuclei (AGN) feedback, metal mixing and diffusion, etc. Because the processes which give rise to this distribution are complex, there is almost certainly some structure in the stellar abundance distribution for every galaxy. However, it has only been definitively measured in the Milky Way, \edit1{with conflicting claims of detection \citep{2023ApJ...956L..14K} and non-detection \citep{2024IAUS..377..115N} in M31.}

% Why alpha-elements are important
The distribution of elemental abundances is a high dimensional space \citep[e.g., 32 elements in][]{2024ApJ...961L..41J}. However, this space is highly degenerate, and so the effective number of dimensions is much smaller -- even possibly compressed to just \FeH{} and age \citep{2019ApJ...883..177N}. Two elements have received particular interest - Fe and elements produced by the $\alpha$-process. Type Ia and Type II supernovae are the main contributors of Fe and $\alpha$ enrichment. Fe is broadly produced in both types, and so its abundance is a proxy for the total metallicity of a star. On the other hand, $\alpha$-elements are mainly produced in Type II supernovae. The ratio of $\alpha$-elements to Fe (\alphaFe{}) is then a measure of the relative contributions of Type Ia and II SNe to the enrichment of a parcel of gas, \edit1{which typically declines with time \citep{1979ApJ...229.1046T,1986A&A...154..279M}}. It has therefore become common to compress the high-dimensional abundance space to the two dimensional \alphaFe{}-\FeH{} plane.

% Observed chemical bimodality in the Milky Way
\edit1{These abundances can be used to decompose the Milky Way's disk, which has a long history dating back to the work of \citet{1983MNRAS.202.1025G}, who noted that the vertical distribution of stellar altitudes is well-fit by a double exponential. This led naturally to a ``thin'' and ``thick'' disk, whose membership can be reasonably determined through kinematics \citep[e.g.][]{2003A&A...410..527B}. It was quickly realized that the thick disk is more $\alpha$-enhanced than the thin disk \citep{1996ASPC...92..307G,1998A&A...338..161F}.}

\edit1{Later studies showed that the disk could be decomposed into high- and low-$\alpha$ sequences without kinematic selection \citep{2011A&A...535L..11A,2012A&A...545A..32A}.\footnote{\citet{2003A&A...410..527B} briefly noted that the thin and thick disk seemed to not overlap in chemistry.} The high-$\alpha$ sequence is older, more centrally compact, and more vertically extended than the low-$\alpha$ sequence \citep{2013A&A...560A.109H,2024IAUS..377..115N}. Although the thick disk is more $\alpha$-enhanced than the thin disk, it is not immediately obvious that the chemical and kinematic separations arise from the same physical process \citep[or that they even exist, see][]{2012ApJ...751..131B}.}

% Different explanations of bimodality
Naturally, many different processes that could lead to structure in the abundance plane have been discussed in the literature. An early explanation of the bimodality is \edit1{based on} the two-phase gas infall model \citep{1997ApJ...477..765C,2009IAUS..254..191C,2017MNRAS.472.3637G,2019A&A...623A..60S}. In this model, the thick disk first forms rapidly from an initial infall of gas. Because the typical SFR is high, these stars are $\alpha$-enhanced. In some variants, star formation halts completely before a second supply of pristine gas falls into the Galaxy \citep[][and references therein]{2024arXiv240511025S}. This dilutes the gas supply from which the thin disk forms more gradually, \edit1{creating a loop feature in the abundance plane}. The thin disk is then more $\alpha$-poor because its associated SFR is lower, \edit1{and in certain scenarios two chemically distinct disks are formed.}

@@

Next, we turn to models which argue the bimodality results from some external influence. Early arguments were made that both the $\alpha$-enhancement of the disk and the thickening of the disk can result from gas-rich mergers \citep{2004ApJ...612..894B,2005ApJ...630..298B,2007ApJ...658...60B,2010MNRAS.402.1489R}.\footnote{See also \citet{2009MNRAS.400.1347C} for an argument invoking semi-analytic models.} These mergers lead to an enhanced SFR which leads to the $\alpha$-enhancement of the thick disk, \edit1{with \citet{2015A&A...578A..87S} being the first to attempt to explain abundance substructure with a merger.}

@@

The merger explanation of the bimodality is highly synergistic with our picture of the hierarchical assembly of the stellar halo \citep{2005ApJ...635..931B}. Indeed, there is strong evidence that the Milky Way underwent a significant merger with the so-called Gaia-Sausage-Enceladus satellite \citep[GSE;][]{2018MNRAS.478..611B,2018Natur.563...85H,2020ApJ...901...48N}. This merger is thought to have occurred $\sim8-10\,\Gyr$ ago \citep[see also][]{2020ApJ...897L..18B}. \edit1{A merger origin of the abundance bimodality is also attractive because it can simultaneously explain the origin of the kinematic thin and thick disk \citep{1985AJ.....90.2015G,1986ApJ...309..472Q,1993ApJ...403...74Q}.}

@@

\edit1{In this work, we propose that a brief $\sim300\,\Myr$ interruption in the formation of stars at a given metallicity can lead to the formation of an $\alpha$-abundance bimodality at that metallicity. It is not common to study the star formation rate at a specific metallicity, but dividing the stellar population into narrow abundance populations can be a powerful tool \citep[e.g.][]{2012ApJ...753..148B,2012ApJ...751..131B}.}

\edit1{Our proposal assumes that the gas phase's \alphaFe{} is declining sufficiently rapidly at the time of the interruption. Using a set of idealized simulations which mimic the $z\sim2$ merger between the Milky Way and GSE, we show that such a merger can drive the formation of this gap and thus the bimodality. While we demonstrate this mechanism in the context of a merger scenario, it is important to note that our proposal does not inherently require a merger to induce this metallicity-dependent star formation gap. This scenario predicts a $\sim300\,\Myr$ gap in stellar ages at metallicities where the bimodality exists ($\FeH\lesssim-0.2$).}

In Section~\ref{sec:methods}, we describe our setup. \edit1{In Section~\ref{sec:results}, we present in detail the main results of two example simulations before expanding our results to the full suite.} In Section~\ref{sec:discussion}, we discuss and interpret our results, as well as connections to previous and future work, before concluding in Section~\ref{sec:conclusion}. Throughout this work we refer to the standard native time unit $\kpc/\left(\kms\right)$ as \Gyr{} for convenience.

@@

We use a modified version of the \texttt{MakeNewDisk} variant described in \citet{2023MNRAS.tmp.2070B}. In isolation, each of the central and satellite galaxies are a compound halo setup, with a \citet{1990ApJ...356..359H} dark matter halo and a gaseous halo with a $\beta$-profile:
\begin{equation*}
\rho = \rho_0 \left[1 + \left(\frac{r}{r_c}\right)^2\right]^{-\frac{3\beta}{2}}
\end{equation*}
The total mass within the virial radius is kept fixed, and the mass of the dark matter halo and central density of the gaseous halo are chosen to satisfy a given baryon fraction $f_b$ \edit1{within the virial radius}. The dark matter halo is initialized to be in gravitational equilibrium with the total potential. The gaseous halo is in gravito-hydrostatic equilibrium, where the temperature is allowed to vary as a function of radius. The azimuthal velocity of the gaseous halo is given as a fraction of the circular velocity. There is no initial stellar disk or bulge, and the gas is initially metal-free. Thus, all star particles and metals are formed self-consistently.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{orbit.pdf}
    \caption{The orbits of the bimodal (blue) and unimodal (orange) simulations. The upper and middle panels show the orbit in the $x$-$y$ and $x$-$z$ planes, respectively. The bottom panel shows the separation distance as a function of time. The orbit begins retrograde but then radializes after the first pericentric passage. The satellite then coalesces quickly after the second pericentric passage, after $\sim2\,\Gyr$ of evolution. \edit1{A blue dashed line is shown for the bimodal simulation at $2.8\,\Gyr$, the time of a star formation gap at $\FeH=0$ (see Figure~\ref{fig:before_after}). An orange dashed line is shown for the unimodal simulation (which has no gap) at the same time after its second pericentric passage ($2.3\,\Gyr$).}}
    \label{fig:orbit}
\end{figure}

@@

Some of the simulations in this orbital grid resulted in bimodal abundance distributions, while some had little to no structure in the abundance distribution plane. \edit1{We will first study two representative simulations in detail chosen based on their structure in the abundance plane as shown in Figure~\ref{fig:fig1}, one which we refer to as bimodal and one as unimodal. For the bimodal simulation, we chose the simulation with $R_0=129\,\kpc$, $V_0=142\,\kms$, and $\eta=0.4$. For the unimodal simulation, the parameters are the same except that $V_0=116\,\kms$. These simulations will later be identified as having the highest and second lowest bimodality score $\mathcal{B}$. We will then examine the full simulation suite, and show the detailed abundance plane for the full suite in Appendix~\ref{app:allmerge}.}

We show the bimodal and unimodal simulations' orbits in Figure~\ref{fig:orbit}. \edit1{We use a shrinking spheres center of mass method to identify the centers of the central and satellite galaxy \citep[e.g.,][]{2003MNRAS.338...14P}.\footnote{The position of the minimum potential particle in each substructure identified by \texttt{SUBFIND} is used as the starting guess, and we use an initial/final radius and step factor of $10\,\kpc$, $5\,\kpc$, and $0.9$, respectively.}} The upper and middle panels show the orbits in the $x$-$y$ and $x$-$z$ planes, respectively. The lower panel shows the separation distance as a function of time. The orbit is initially retrograde, but quickly radializes after the first pericentric passage. Coalescence occurs rapidly after the second pericentric passage at $\sim2\,\Gyr$, and \texttt{SUBFIND} ceases to recognize the satellite as a separate subhalo.

\subsection{Feedback and Enrichment Model}\label{ssec:gfm}
Our feedback model is \edit1{a variant of the Illustris TNG model} \citep{2013MNRAS.436.3031V,2017MNRAS.465.3291W,2018MNRAS.473.4077P}. In this model, gravity and magnetohydrodynamics are solved using a \citet{1986Natur.324..446B} tree coupled to a second order finite volume fluid solver in AREPO \citep{2010MNRAS.401..791S,2016MNRAS.455.1134P}. Stellar feedback is included through a subgrid wind particle model \citep{2003MNRAS.339..289S}. AGN feedback follows a dual kinetic and thermal mode for low- and high-accretion rates \citep{2017MNRAS.465.3291W}, though in our setup the AGN is only ever in the high-accretion mode. The central galaxy is seeded with a black hole with the typical seed mass ($8\times10^5\,\Msun$).

@@

In this model, star particles synthesize elements through three different channels for which we cite the relevant yield tables: SNe Ia \citep{1997NuPhA.621..467N}, SNe II \citep{1998A&A...334..505P,2006ApJ...653.1145K}, and AGB stars \citep{2010MNRAS.403.1413K,2014MNRAS.437..195D,2014ApJ...797...44F}. Each star particle, \edit1{which is modeled as a simple stellar population}, continuously injects metals into its surroundings in the following sequence\footnote{The kinetic/thermal feedback component is handled through the wind generation, which is completely separate in this model.}:
\begin{enumerate}
    \item $t\lesssim10\,\Myr$: no metal injection as the first supernova ($M\sim100\,\Msun$) has not gone off
    \item $10\,\Myr \lesssim t \lesssim 40\,\Myr$: metal injection as $8\,\Msun<M<100\,\Msun$ stars die as Type II SNe
    \item $t\gtrsim40\,\Myr$: metal injection from Type Ia SNe and AGB stars
\end{enumerate}
There are a few things to note about this model: (1) The exact timings are metallicity-dependent. (2) The \MgFe{} of ejected gas from Type II SNe is mass/time-dependent, with more massive stars contributing more Mg than less massive stars. (3) In a Hubble time, type II SNe contribute the vast majority of Mg ($\sim10\times$ AGB and $\sim100\times$ Type Ia SNe). Type Ia and Type II SNe contribute approximately equal amounts of Fe (each $\sim3\times$ AGB). See Figure~1 from \citet{2018MNRAS.473.4077P}. (4) The number of Type Ia SNe is greater for a younger stellar population, with a power law relationship $\propto \left(t/\tau_8\right)^{-1.12}$, where $\tau_8=40\,\Myr$ is the lifetime of an $8\,\Msun$ star.

@@

We \edit1{applied quality cuts and restricted our sample to giants}, requiring:
\begin{itemize}[noitemsep]
    \item $\textrm{SNR} > 200$,
    \item $\textrm{VSCATTER} < 1\,\kms$,
    \item STARFLAG not set,
    \item $\varpi/\sigma_{\varpi} > 1$,
    \item $\log{g} < 3.5$,
    \item $\sigma_{\log{g}} < 0.2$,
\end{itemize}
where $\varpi$ is the parallax. We use the parallax, proper motion, and radial velocity from Gaia EDR3 \citep{2016AA...595A...1G,2021AA...649A...1G,2021AA...649A...2L,2021AA...653A.160S}.

\edit1{We next make a solar neighborhood selection of stars based on their angular momenta.} We assume the solar radius and azimuthal velocity are $R_0=8\,\kpc$ and $V_0=220\,\kms$ \citep{2016ARA&A..54..529B}, and select stars which have $L_z$ within $10\%$ of the solar angular momentum. We further require that $\left|z\right| < 3\,\kpc$. As is typically done, we use \FeH{} as an indicator of the total metallicity of a star. We use Mg alone as a representative of the $\alpha$-elements.

\edit1{For stellar ages, we used the APOKASC-3 catalog \citep{2025ApJS..276...69P}. This catalog uses a combination of APOGEE spectroscopic parameters and \textit{Kepler} time series photometry to compute astroseismic ages. Using only stars with $25\%$ age uncertainties (taken as the maximum of the upper and lower uncertainty), we cross-match this catalog to our larger sample from ASPCAP which results in a sample of 2525 stars.}

@@

\subsection{Bimodality Score}\label{ssec:bim_score}
\edit1{Given the modest size of our suite, some method for scoring the degree of bimodality for a given 1D distribution is desirable. Tests of whether a distribution is bimodal or unimodal exist -- e.g., the Hartigan dip test \citep{10.1214/aos/1176346577}, but they lack the ability to rank order based on ``bimodalitiness.'' In order to do this, we fit a given distribution of \alphaFe{} as a two-component Gaussian mixture model. The bimodality score $\mathcal{B}$ is then computed as
\begin{equation}\label{eq:bimscore}
\mathcal{B} = \frac{\left|\mu_1-\mu_2\right|}{\sqrt{\sigma_1^2-\sigma_2^2}}\times w(A_2, t, k)\textrm{,}
\end{equation}
where $(\mu_1, \sigma_1)$ is the mean and standard deviation of the higher weighted mode, $(\mu_2, \sigma_2)$ likewise for the less weighted mode, $A_2$ is the amplitude of the less weighted mode, and $w$ is a penalty function with parameters $t$ and $k$ given by
\begin{equation}\label{eq:bimscore_penalty}
w(A, t, k) = \left[1+e^{-k\left(A-t\right)}\right]^{-1}\text{.}
\end{equation}
We take $t$ and $k$ to be $0.1$ and $20$, respectively.}

\edit1{This metric effectively measures the distance between two modes normalized by their variances, with a penalty when one mode is not highly weighted. These choices were made in order for the score to match by eye which distributions appear to have one as opposed to two peaks, and was found to empirically be more reliable than other statistics like peak-to-trough ratios or mode overlap. As demonstrated in Section~\ref{ssec:full_sim_suite}, a score of $\mathcal{B}>2.25$ appears to select the bimodal populations.}

@@

\subsection{Build up of the Abundance Plane}\label{ssec:abundplane_build}
Next, we examine the build up of the abundance plane. \edit1{In the left panel of Figure~\ref{fig:before_after}, we show the metallicity-dependent star formation rate of the bimodal simulation using star particles in the solar neighborhood at the end of the simulation within a $0.1\,\dex$ bin centered at $\FeH=0$. There is a clear dip in the SFR at $\sim2.8\,\Gyr$, which we indicate with a vertical line.}

\edit1{In the center left panel, we show the abundance plane distribution from the bimodal simulation for all stars in the solar neighborhood, replicating Figure~\ref{fig:fig1}. A dashed line at $-0.1\times\FeH+0.3$, chosen by eye, is plotted to demarcate the high- and low-$\alpha$ sequences. In the center right and right panels, we show the distribution of star particles which form before and after the dip in the metallicity-dependent SFR at $t=2.8\,\Gyr$, respectively. The vast majority of the high-$\alpha$ sequence forms before the dip, while most of the low-$\alpha$ sequence forms afterward.}

>>>>

\begin{abstract}
    The elemental abundance distribution of stars encodes the history of the gas-phase abundance in the Milky Way. Without a large, unbiased sample of highly precise stellar ages, the exact timing and nature of this history must be \textit{inferred} from the abundances. In the two-dimensional plane of \alphaFe{}-\FeH{}, it is now clear that two separate populations exist -- the low-$\alpha$ and high-$\alpha$ sequences. We propose that a brief ($\sim300\,\Myr$) halt in star formation within a narrow metallicity bin can lead to a bimodal \alphaFe{} distribution at that metallicity, assuming a rapidly declining gas phase \alphaFe{}. Using simulations of an idealized setup of a high-$z$ galaxy merger, we show that the merger with the Gaia-Sausage-Enceladus satellite at $z\sim2$ is one possible way to trigger such a gap in the Milky Way. This mechanism may also operate in non-merger scenarios. We predict a $\sim300\,\Myr$ gap in stellar ages at a fixed \FeH{} where the $\alpha$-bimodality is prominent ($\FeH\lesssim-0.2$).
\end{abstract}

@@

\section{Introduction} \label{sec:intro}
% What leads to distribution of heavy elements in a galaxy
Many elements heavier than hydrogen are produced through nuclear fusion in compact objects such as supernovae, dying low mass stars, and neutron star-neutron star mergers \citep[e.g.][]{2023A&ARv..31....1A}. By necessity, stars inherit the constitutive properties of the gas from which they formed. Moreover, the surface abundance of most elements for most stars do not change over most of their lifetime. By analyzing the surface abundances of stars, we can reconstruct the historical gas-phase composition of a galaxy.

The enrichment of the gas-phase of a galaxy is determined by a complicated combination of physical processes - stellar evolution and supernovae, gas accretion, mergers, gas outflows from stellar and active galactic nuclei (AGN) feedback, metal mixing and diffusion, etc. Because the processes which give rise to this distribution are complex, there is almost certainly some structure in the stellar abundance distribution for every galaxy. However, it has only been definitively measured in the Milky Way, with conflicting claims of detection \citep{2023ApJ...956L..14K} and non-detection \citep{2024IAUS..377..115N} in M31.

% Why alpha-elements are important
The distribution of elemental abundances is a high dimensional space \citep[e.g., 32 elements in][]{2024ApJ...961L..41J}. However, this space is highly degenerate, and so the effective number of dimensions is much smaller -- even possibly compressed to just \FeH{} and age \citep{2019ApJ...883..177N}. Two elements have received particular interest - Fe and elements produced by the $\alpha$-process. Type Ia and Type II supernovae are the main contributors of Fe and $\alpha$ enrichment. Fe is broadly produced in both types, and so its abundance is a proxy for the total metallicity of a star. On the other hand, $\alpha$-elements are mainly produced in Type II supernovae. The ratio of $\alpha$-elements to Fe (\alphaFe{}) is then a measure of the relative contributions of Type Ia and II SNe to the enrichment of a parcel of gas, which typically declines with time \citep{1979ApJ...229.1046T,1986A&A...154..279M}. It has therefore become common to compress the high-dimensional abundance space to the two dimensional \alphaFe{}-\FeH{} plane.

% Observed chemical bimodality in the Milky Way
These abundances can be used to decompose the Milky Way's disk, which has a long history dating back to the work of \citet{1983MNRAS.202.1025G}, who noted that the vertical distribution of stellar altitudes is well-fit by a double exponential. This led naturally to a ``thin'' and ``thick'' disk, whose membership can be reasonably determined through kinematics \citep[e.g.][]{2003A&A...410..527B}. It was quickly realized that the thick disk is more $\alpha$-enhanced than the thin disk \citep{1996ASPC...92..307G,1998A&A...338..161F}.

Later studies showed that the disk could be decomposed into high- and low-$\alpha$ sequences without kinematic selection \citep{2011A&A...535L..11A,2012A&A...545A..32A}.\footnote{\citet{2003A&A...410..527B} briefly noted that the thin and thick disk seemed to not overlap in chemistry.} The high-$\alpha$ sequence is older, more centrally compact, and more vertically extended than the low-$\alpha$ sequence \citep{2013A&A...560A.109H,2024IAUS..377..115N}. Although the thick disk is more $\alpha$-enhanced than the thin disk, it is not immediately obvious that the chemical and kinematic separations arise from the same physical process \citep[or that they even exist, see][]{2012ApJ...751..131B}.

% Different explanations of bimodality
Naturally, many different processes that could lead to structure in the abundance plane have been discussed in the literature. An early explanation of the bimodality is based on the two-phase gas infall model \citep{1997ApJ...477..765C,2009IAUS..254..191C,2017MNRAS.472.3637G,2019A&A...623A..60S}. In this model, the thick disk first forms rapidly from an initial infall of gas. Because the typical SFR is high, these stars are $\alpha$-enhanced. In some variants, star formation halts completely before a second supply of pristine gas falls into the Galaxy \citep[][and references therein]{2024arXiv240511025S}. This dilutes the gas supply from which the thin disk forms more gradually, creating a loop feature in the abundance plane. The thin disk is then more $\alpha$-poor because its associated SFR is lower, and in certain scenarios two chemically distinct disks are formed.

@@

Next, we turn to models which argue the bimodality results from some external influence. Early arguments were made that both the $\alpha$-enhancement of the disk and the thickening of the disk can result from gas-rich mergers \citep{2004ApJ...612..894B,2005ApJ...630..298B,2007ApJ...658...60B,2010MNRAS.402.1489R}.\footnote{See also \citet{2009MNRAS.400.1347C} for an argument invoking semi-analytic models.} These mergers lead to an enhanced SFR which leads to the $\alpha$-enhancement of the thick disk, with \citet{2015A&A...578A..87S} being the first to attempt to explain abundance substructure with a merger.

@@

The merger explanation of the bimodality is highly synergistic with our picture of the hierarchical assembly of the stellar halo \citep{2005ApJ...635..931B}. Indeed, there is strong evidence that the Milky Way underwent a significant merger with the so-called Gaia-Sausage-Enceladus satellite \citep[GSE;][]{2018MNRAS.478..611B,2018Natur.563...85H,2020ApJ...901...48N}. This merger is thought to have occurred $\sim8-10\,\Gyr$ ago \citep[see also][]{2020ApJ...897L..18B}. A merger origin of the abundance bimodality is also attractive because it can simultaneously explain the origin of the kinematic thin and thick disk \citep{1985AJ.....90.2015G,1986ApJ...309..472Q,1993ApJ...403...74Q}.

@@

In this work, we propose that a brief $\sim300\,\Myr$ interruption in the formation of stars at a given metallicity can lead to the formation of an $\alpha$-abundance bimodality at that metallicity. It is not common to study the star formation rate at a specific metallicity, but dividing the stellar population into narrow abundance populations can be a powerful tool \citep[e.g.][]{2012ApJ...753..148B,2012ApJ...751..131B}.

Our proposal assumes that the gas phase's \alphaFe{} is declining sufficiently rapidly at the time of the interruption. Using a set of idealized simulations which mimic the $z\sim2$ merger between the Milky Way and GSE, we show that such a merger can drive the formation of this gap and thus the bimodality. While we demonstrate this mechanism in the context of a merger scenario, it is important to note that our proposal does not inherently require a merger to induce this metallicity-dependent star formation gap. This scenario predicts a $\sim300\,\Myr$ gap in stellar ages at metallicities where the bimodality exists ($\FeH\lesssim-0.2$).

In Section~\ref{sec:methods}, we describe our setup. In Section~\ref{sec:results}, we present in detail the main results of two example simulations before expanding our results to the full suite. In Section~\ref{sec:discussion}, we discuss and interpret our results, as well as connections to previous and future work, before concluding in Section~\ref{sec:conclusion}. Throughout this work we refer to the standard native time unit $\kpc/\left(\kms\right)$ as \Gyr{} for convenience.

@@

We use a modified version of the \texttt{MakeNewDisk} variant described in \citet{2023MNRAS.tmp.2070B}. In isolation, each of the central and satellite galaxies are a compound halo setup, with a \citet{1990ApJ...356..359H} dark matter halo and a gaseous halo with a $\beta$-profile:
\begin{equation*}
\rho = \rho_0 \left[1 + \left(\frac{r}{r_c}\right)^2\right]^{-\frac{3\beta}{2}}
\end{equation*}
The total mass within the virial radius is kept fixed, and the mass of the dark matter halo and central density of the gaseous halo are chosen to satisfy a given baryon fraction $f_b$ within the virial radius. The dark matter halo is initialized to be in gravitational equilibrium with the total potential. The gaseous halo is in gravito-hydrostatic equilibrium, where the temperature is allowed to vary as a function of radius. The azimuthal velocity of the gaseous halo is given as a fraction of the circular velocity. There is no initial stellar disk or bulge, and the gas is initially metal-free. Thus, all star particles and metals are formed self-consistently.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{orbit.pdf}
    \caption{The orbits of the bimodal (blue) and unimodal (orange) simulations. The upper and middle panels show the orbit in the $x$-$y$ and $x$-$z$ planes, respectively. The bottom panel shows the separation distance as a function of time. The orbit begins retrograde but then radializes after the first pericentric passage. The satellite then coalesces quickly after the second pericentric passage, after $\sim2\,\Gyr$ of evolution. A blue dashed line is shown for the bimodal simulation at $2.8\,\Gyr$, the time of a star formation gap at $\FeH=0$ (see Figure~\ref{fig:before_after}). An orange dashed line is shown for the unimodal simulation (which has no gap) at the same time after its second pericentric passage ($2.3\,\Gyr$).}
    \label{fig:orbit}
\end{figure}

@@

Some of the simulations in this orbital grid resulted in bimodal abundance distributions, while some had little to no structure in the abundance distribution plane. We will first study two representative simulations in detail chosen based on their structure in the abundance plane as shown in Figure~\ref{fig:fig1}, one which we refer to as bimodal and one as unimodal. For the bimodal simulation, we chose the simulation with $R_0=129\,\kpc$, $V_0=142\,\kms$, and $\eta=0.4$. For the unimodal simulation, the parameters are the same except that $V_0=116\,\kms$. These simulations will later be identified as having the highest and second lowest bimodality score $\mathcal{B}$. We will then examine the full simulation suite, and show the detailed abundance plane for the full suite in Appendix~\ref{app:allmerge}.

We show the bimodal and unimodal simulations' orbits in Figure~\ref{fig:orbit}. We use a shrinking spheres center of mass method to identify the centers of the central and satellite galaxy \citep[e.g.,][]{2003MNRAS.338...14P}.\footnote{The position of the minimum potential particle in each substructure identified by \texttt{SUBFIND} is used as the starting guess, and we use an initial/final radius and step factor of $10\,\kpc$, $5\,\kpc$, and $0.9$, respectively.} The upper and middle panels show the orbits in the $x$-$y$ and $x$-$z$ planes, respectively. The lower panel shows the separation distance as a function of time. The orbit is initially retrograde, but quickly radializes after the first pericentric passage. Coalescence occurs rapidly after the second pericentric passage at $\sim2\,\Gyr$, and \texttt{SUBFIND} ceases to recognize the satellite as a separate subhalo.

\subsection{Feedback and Enrichment Model}\label{ssec:gfm}
Our feedback model is a variant of the Illustris TNG model \citep{2013MNRAS.436.3031V,2017MNRAS.465.3291W,2018MNRAS.473.4077P}. In this model, gravity and magnetohydrodynamics are solved using a \citet{1986Natur.324..446B} tree coupled to a second order finite volume fluid solver in AREPO \citep{2010MNRAS.401..791S,2016MNRAS.455.1134P}. Stellar feedback is included through a subgrid wind particle model \citep{2003MNRAS.339..289S}. AGN feedback follows a dual kinetic and thermal mode for low- and high-accretion rates \citep{2017MNRAS.465.3291W}, though in our setup the AGN is only ever in the high-accretion mode. The central galaxy is seeded with a black hole with the typical seed mass ($8\times10^5\,\Msun$).

@@

In this model, star particles synthesize elements through three different channels for which we cite the relevant yield tables: SNe Ia \citep{1997NuPhA.621..467N}, SNe II \citep{1998A&A...334..505P,2006ApJ...653.1145K}, and AGB stars \citep{2010MNRAS.403.1413K,2014MNRAS.437..195D,2014ApJ...797...44F}. Each star particle, which is modeled as a simple stellar population, continuously injects metals into its surroundings in the following sequence\footnote{The kinetic/thermal feedback component is handled through the wind generation, which is completely separate in this model.}:
\begin{enumerate}
    \item $t\lesssim10\,\Myr$: no metal injection as the first supernova ($M\sim100\,\Msun$) has not gone off
    \item $10\,\Myr \lesssim t \lesssim 40\,\Myr$: metal injection as $8\,\Msun<M<100\,\Msun$ stars die as Type II SNe
    \item $t\gtrsim40\,\Myr$: metal injection from Type Ia SNe and AGB stars
\end{enumerate}
There are a few things to note about this model: (1) The exact timings are metallicity-dependent. (2) The \MgFe{} of ejected gas from Type II SNe is mass/time-dependent, with more massive stars contributing more Mg than less massive stars. (3) In a Hubble time, type II SNe contribute the vast majority of Mg ($\sim10\times$ AGB and $\sim100\times$ Type Ia SNe). Type Ia and Type II SNe contribute approximately equal amounts of Fe (each $\sim3\times$ AGB). See Figure~1 from \citet{2018MNRAS.473.4077P}. (4) The number of Type Ia SNe is greater for a younger stellar population, with a power law relationship $\propto \left(t/\tau_8\right)^{-1.12}$, where $\tau_8=40\,\Myr$ is the lifetime of an $8\,\Msun$ star.

@@

We applied quality cuts and restricted our sample to giants, requiring:
\begin{itemize}[noitemsep]
    \item $\textrm{SNR} > 200$,
    \item $\textrm{VSCATTER} < 1\,\kms$,
    \item STARFLAG not set,
    \item $\varpi/\sigma_{\varpi} > 1$,
    \item $\log{g} < 3.5$,
    \item $\sigma_{\log{g}} < 0.2$,
\end{itemize}
where $\varpi$ is the parallax. We use the parallax, proper motion, and radial velocity from Gaia EDR3 \citep{2016AA...595A...1G,2021AA...649A...1G,2021AA...649A...2L,2021AA...653A.160S}.

We next make a solar neighborhood selection of stars based on their angular momenta. We assume the solar radius and azimuthal velocity are $R_0=8\,\kpc$ and $V_0=220\,\kms$ \citep{2016ARA&A..54..529B}, and select stars which have $L_z$ within $10\%$ of the solar angular momentum. We further require that $\left|z\right| < 3\,\kpc$. As is typically done, we use \FeH{} as an indicator of the total metallicity of a star. We use Mg alone as a representative of the $\alpha$-elements.

For stellar ages, we used the APOKASC-3 catalog \citep{2025ApJS..276...69P}. This catalog uses a combination of APOGEE spectroscopic parameters and \textit{Kepler} time series photometry to compute astroseismic ages. Using only stars with $25\%$ age uncertainties (taken as the maximum of the upper and lower uncertainty), we cross-match this catalog to our larger sample from ASPCAP which results in a sample of 2525 stars.

@@

\subsection{Bimodality Score}\label{ssec:bim_score}
Given the modest size of our suite, some method for scoring the degree of bimodality for a given 1D distribution is desirable. Tests of whether a distribution is bimodal or unimodal exist -- e.g., the Hartigan dip test \citep{10.1214/aos/1176346577}, but they lack the ability to rank order based on ``bimodalitiness.'' In order to do this, we fit a given distribution of \alphaFe{} as a two-component Gaussian mixture model. The bimodality score $\mathcal{B}$ is then computed as
\begin{equation}\label{eq:bimscore}
\mathcal{B} = \frac{\left|\mu_1-\mu_2\right|}{\sqrt{\sigma_1^2-\sigma_2^2}}\times w(A_2, t, k)\textrm{,}
\end{equation}
where $(\mu_1, \sigma_1)$ is the mean and standard deviation of the higher weighted mode, $(\mu_2, \sigma_2)$ likewise for the less weighted mode, $A_2$ is the amplitude of the less weighted mode, and $w$ is a penalty function with parameters $t$ and $k$ given by
\begin{equation}\label{eq:bimscore_penalty}
w(A, t, k) = \left[1+e^{-k\left(A-t\right)}\right]^{-1}\text{.}
\end{equation}
We take $t$ and $k$ to be $0.1$ and $20$, respectively.

This metric effectively measures the distance between two modes normalized by their variances, with a penalty when one mode is not highly weighted. These choices were made in order for the score to match by eye which distributions appear to have one as opposed to two peaks, and was found to empirically be more reliable than other statistics like peak-to-trough ratios or mode overlap. As demonstrated in Section~\ref{ssec:full_sim_suite}, a score of $\mathcal{B}>2.25$ appears to select the bimodal populations.

@@

\subsection{Build up of the Abundance Plane}\label{ssec:abundplane_build}
Next, we examine the build up of the abundance plane. In the left panel of Figure~\ref{fig:before_after}, we show the metallicity-dependent star formation rate of the bimodal simulation using star particles in the solar neighborhood at the end of the simulation within a $0.1\,\dex$ bin centered at $\FeH=0$. There is a clear dip in the SFR at $\sim2.8\,\Gyr$, which we indicate with a vertical line.

In the center left panel, we show the abundance plane distribution from the bimodal simulation for all stars in the solar neighborhood, replicating Figure~\ref{fig:fig1}. A dashed line at $-0.1\times\FeH+0.3$, chosen by eye, is plotted to demarcate the high- and low-$\alpha$ sequences. In the center right and right panels, we show the distribution of star particles which form before and after the dip in the metallicity-dependent SFR at $t=2.8\,\Gyr$, respectively. The vast majority of the high-$\alpha$ sequence forms before the dip, while most of the low-$\alpha$ sequence forms afterward.
